---
title: "R Notebook"
output: html_notebook
---

name.of.function <- function(arg1, arg2) {
  statements
  return(something)
}

name (can be almost anything)
arguments (i.e input to your funciton)
body (where the work gets done)

```{r}
add <- function(x, y=1){
  #sum the input x and y
  x+y
}
```
1. name : add
2. arguments: here x and y
3. body: will return result of last statement

```{r}
add <- function(x, y=1){
  #sum the input x and y
  x+y
}
add(x=4)
add(x=4, y=10)
add(4,20)
add(c(1,2,5,10),y=10)
```

having to write the same thing multiple times -> write a function
why you would write a funciton:
- makes the purpose of the code more clear
- reduce mistakes from copy and paste
- makes updating code easier
- reduce duplication and facilitate re-use

start with a working code snippet, simplify, reduce calculation duplication
consider original code:
```{r} 
df$a <- (df$a-min(df$a))/max(df$a)-min(df$a))
```

simplify to work with a generic vector named "x"
```{r}
xmin <- min(x)
x <- (x-min(x))/(max(x)-min(x))

```
range function also works for min and max range()

turn it into function/put into body
put it into console i guess
```{r}
rescale <- function(x) {
  rng <- range(x)
  (x-rng[1])/(rng[2]-rng[1])
}
```
test on a small example where you know the answer
```{r}
rescale(1:10)
rescale(c(1,4,10,20))
rescale(c(1, NA, 4, 10,20))
```

```{r}
range(c(1,NA,4,10,20), na.rm=TRUE)
```

fix the NA issue: 

```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm=TRUE)
  (x-rng[1])/(rng[2]-rng[1])
}
rescale01(c(1,NA, 4, 10,20))
```

can source other R scripts/functions with source() thing, but gotta save it as R script