---
title: "Lab 08"
author: "Cody Hargadon A15479290"
date: "10/21/2021"
output: html_document
---

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

>Q1. Number of rows and columns: 

number of rows:
```{r}
nrow(x)
```
number of columns:
```{r}
ncol(x)
```

checking the data
```{r}

#View(x)
dim(x)
```

fixing row names
if you run this twice, you can lose things so probably do it the second way
```{r}
#rownames(x) <- x[,1]
#x<- x[,-1]
#head(x)
#dim(x)
```

alternative way to do this:
```{r}
x <- read.csv(url, row.names=1)
head(x)
```

>Q2.  Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

the first option you can only run once, so if you happen to run the entire code again you can overide things and lose columns. the second way to do it makes more sense as you correctly change it at the same time as reading the data set. 


#Spotting major differences and trends

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

>Q3. Changing what optional argument in the barplot() function results in the following graph?

changing "beside=T" to "beside=F" changes the graph to the following
```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

>Q5. Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(10), pch=16)
```

if a point lies on the diagonal for a given plot it means the country being plotted against the others has a high similarity in that category to that specific country.


#PCA! 
the main function in base R for PCA is 'prcomp'
this wants the transpose of our data

>Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?


```{r}
t(x)
pca <- prcomp( t(x) )
summary(pca)
```
```{r}
attributes(pca)
plot(pca$x[,1], pca$x[,2])
```


>Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))

```

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
colors <- c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x), col= colors)
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```
or
```{r}
z <- summary(pca)
z$importance
```


```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

#Variable loading
```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

>Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

PC2 tells us the specific variation that PC1 does not. Fresh potatoes and soft drinks feature prodominantely. 

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```


#Biplots

```{r}
biplot(pca)
```

#PCA of RNA-seq data

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```
>Q10: How many genes and samples are in this data set?

```{r}
nrow(rna.data)
ncol(rna.data)
```
There are 100 rows (genes) and 10 columns (samples)


