---
title: "mini project"
author: "Cody"
date: "10/26/2021"
output: word_document
---

```{r}
setwd("~/Desktop")
fna.data <- read.csv("WisconsinCancer.csv")
#fna.data
wisc.df <- data.frame(fna.data, row.names=1)
#wisc.df
wisc.data <- wisc.df[,-1]
#wisc.data
diagnosis <- as.factor(wisc.df$diagnosis)
```

>Q1. How many observations are in this dataset?

```{r}
nrow(fna.data)
```
There are 569 observations in the data set, 32 observations per patient

>Q2. How many of the observations have a malignant diagnosis?

```{r}
table(diagnosis)
```
There are 212 observations with a malignant diagnosis

>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))
```

There are 10 variables/features in the data set with the suffix "_mean"


#checking col means and standard deviations

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)

#PCA
wisc.pr <- prcomp(wisc.data, scale= TRUE, center=T)
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27% of the variance is captured in the first principle components

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

```{r}
0.4427+0.1897+0.09393
```
PC's 1:3 are required to describe at least 70% of the original variance

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

```{r}
sum(0.4427,0.1897 ,0.09393, 0.06602 ,0.05496, 0.04025 ,0.02251)
```

the first 7 PCs are required to describe at least 90% of the original variance


>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

it is shit. cant understand it because it is chaos

```{r}
biplot(wisc.pr)
```


>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x [, c(1,2)], col = (diagnosis) , xlab = "PC1", ylab = "PC2")
plot(wisc.pr$x[,c(1,3)], col = diagnosis, xlab= "PC1", ylab = "PC3")
```

These plots are much more clear and easy to interpret


>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

concave.points_mean = -0.26085376
```{r}
wisc.pr$rotation[,1]
```


>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

The minimum number of principal components required to explain 80% of variance is 5.

```{r}
sum(0.4427, 0.1897, 0.09393, 0.06602, 0.05496)
summary(wisc.pr)
```


```{r}
library(ggplot2)

df<- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

ggplot(df, aes(PC1, PC2, col= diagnosis)) +
         geom_point()
```


#variance
```{r}
pr.var <- (wisc.pr$sdev)^2
head(pr.var)
pr.var

pve<- pr.var/sum(pr.var)

plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")


barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```




```{r}
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```



#Hierarchical clustering
```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method="complete")
```


>Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

at a height of 20, the clustering model has 4 clusters
```{r}
plot.wisc <- plot(wisc.hclust)
abline(h=4, col="red")
unique(cutree(wisc.hclust, h=20))
```

>Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

  no, the 4 clusters seems to show the best matches.

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters, diagnosis)
```



#5 combining methods
we take the results of our PCA analysis and cluster in this space 'wisc.pr$x"
```{r}
summary(wisc.pr)
```


>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

ward.D2 gives the best results for the data set as it is more defined and informative.
```{r}
wisc.pc.hclust <- hclust(dist(wisc.pr$x[,1:3]), 
                                  method="ward.D2")
plot(wisc.pc.hclust)
abline(h=50, col="red")
```
cut the tree into 2 groups
```{r}
grps <- cutree(wisc.pc.hclust, k=2)
table(grps)
```

>Q15. How well does the newly created model with four clusters separate out the two diagnoses?

  pretty well, its a lot more clear and easier to identify each diagnosis and the false negatives/positives for each. 
  
cross table comparison of diagnosis and my cluster groups
```{r}
table(diagnosis, grps)
```

24: number of false positives
179: called correctly

>Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.



#6 sensitivity/specificity
**accuracy** what proportion did we get correct if we call cluster 1 M and cluster 2 B
```{r}
(333+179)/nrow(wisc.data)
```

**sensitivity** refers to a test’s ability to correctly detect ill patients who do have the condition. In our example here the sensitivity is the total number of samples in the cluster identified as predominantly malignant (cancerous) divided by the total number of known malignant samples. In other words: TP/(TP+FN).
**specificity** relates to a test’s ability to correctly reject healthy patients without a condition. In our example specificity is the proportion of benign (not cancerous) samples in the cluster identified as predominantly benign that are known to be benign. In other words: TN/(TN+FN).

>Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

hierarchal 

sensitivity:
```{r}
179/(179+33)
```

specificity:
```{r}
333/(333+24)
```

>Q18. Which of these new patients should we prioritize for follow up based on your results?

patient 2 should be prioritized for a follow up

```{r}
setwd("~/Desktop")
url <- read.csv("new_samples.csv")
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

